---
title: "Visualizing Numerical Variables in Biomedical Research"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"

format: 
  revealjs:
    #theme: solarized
    css: style.css
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    logo: images/Flame.jpg
    #css: styles.css
    
---

## Lecture Outline


- Distinguishing numeric vs. categorical variables in clinical datasets

- What numeric values represent: physiologic processes, exposure intensity, or outcome magnitude

- Types of numeric variables: continuous (creatinine, systolic BP) vs. discrete (cell counts, days

- How biostatisticians see data before modeling — visual inspection as a form of scientific reasoning

::: footer

:::

## <span style="font-size:80%">Recap: Categorical vs. Numerical</span> {.smaller}

::: columns
::: {.column width="50%"}
**Categorical Variables (Qualitative)**

- Definition: Descriptors of qualities or categories of data

  - Represent labels, names, or groups


- Values: 
  
  - Text 
  - Numbers [i.e.: patient IDs, zip codes]

- Utility: Can count frequencies, mode, or proportions

- Examples:
  - Treatment group: Placebo, Drug A, Drug B


:::

::: {.column width="50%"}
**Numerical Variables (Quantitative)**

- Definition: represent quantities that are measurable and numeric

- Values: Numbers where mathematical operations(addition, subtraction, averaging) make sense

- Types:

  - Discrete: Countable values 

    - Number of hospital visits

  - Continuous: Measurable values that can take on decimals 

    - Blood pressure, cholesterol level

- Utility: allows for a wide range of statistical analyses

- Examples:

  - Serum glucose level (mg/dL)



:::

:::
::: footer
:::

## <span style="font-size:80%">Recap: Categorical vs. Numerical</span> {.smaller}

::: columns
::: {.column width="0%"}



:::

::: {.column width="100%"}
**Numerical Variables (Quantitative)**

- Definition: represent quantities that are measurable and numeric

- Values: Numbers where mathematical operations(addition, subtraction, averaging) make sense

- Types:

  - Discrete: Countable values 

    - Number of hospital visits

  - Continuous: Measurable values that can take on decimals 

    - Blood pressure, cholesterol level

- Utility: allows for a wide range of statistical analyses

- Examples:

  - Serum glucose level (mg/dL)



:::

:::
::: footer
:::



## <span style="font-size:80%">Understanding What Numbers Represent</span> {.smaller}

::: columns
::: {.column width="0%"}



:::

::: {.column width="100%"}
- Numeric variables quantify magnitude, intensity, or duration of biological phenomena

- Each number reflects a measurable process or state:

  - Glucose $\rightarrow$ metabolic control

  - Creatinine $\rightarrow$ renal filtration

  - Ejection fraction $\rightarrow$ cardiac performance

- Biomedical inference begins with understanding what the number means physiologically, not statistically

- Cognitive emphasis: A number’s interpretability depends on context — unit, scale, reference range


:::

:::
::: footer
:::



## <span style="font-size:80%">Why Numeric Data Visualization Matters</span> {.smaller}

::: columns
::: {.column width="0%"}



:::

::: {.column width="100%"}
- Almost every biomedical dataset contains numeric measures that capture biological, clinical, or operational observations

  - Clinical: vital signs, laboratory values, dosage levels

  - Molecular: gene expression counts, protein concentrations

  - Operational: procedure durations, cost per encounter


:::

:::
::: footer
:::


## <span style="font-size:80%">Data Visualization: Seeing meaning in numbers</span> {.smaller}

::: columns
::: {.column width="0%"}



:::

::: {.column width="100%"}
Numeric visualization:

- Supports scientific reasoning through visual comparison: spotting differences in magnitude, rate, or variability across contexts
- Promotes transparency and trust by letting stakeholders visually assess data integrity and trends
- Framing idea: Numeric visualization is visual reasoning with quantities — the foundation for credible biomedical insight


:::

:::
::: footer
:::


## <span style="font-size:80%">Data Visualization: Seeing meaning in numbers</span> {.smaller}

::: columns
::: {.column width="0%"}



:::

::: {.column width="100%"}
Numeric visualization:

- Makes quantities visible — transforming raw numbers into interpretable biomedical meaning
- Serves as an early diagnostic lens to detect outliers, implausible values, or data entry errors
- Allows analysts to see structure, scale, and variation directly — long before statistical modeling is considered

:::

:::
::: footer
:::


## <span style="font-size:80%">Numeric Data in Context</span> {.smaller}

::: columns
::: {.column width="0%"}



:::

::: {.column width="100%"}
- Numeric values have context:

  - Clinical thresholds: normal vs abnormal (e.g., glucose ≥ 126 mg/dL)

  - Temporal frame: before vs after intervention

  - Population frame: adult vs pediatric norms


- Visualize with reference markers or shaded normal ranges when relevant


:::

:::
::: footer
:::


## <span style="font-size:80%">Numeric Data in Context</span> {.smaller}

::: columns
::: {.column width="0%"}



:::

::: {.column width="100%"}
- Annotate with context (mean line, target zone, or standard-of-care threshold)
- Encourage consistency in units and axis scaling across related visuals
- Design goal: viewers should immediately know what “normal” looks like

:::

:::
::: footer
:::







<style>
.tiny-table table { font-size: 0.8em; }        /* adjust as needed (0.8–0.9em) */
.tiny-table figcaption, .tiny-table .table-caption { font-size: 0.75em; }
</style>

## <span style="font-size:80%">Example: Serum Creatinine Distribution</span> {.smaller}

::: columns
::: {.column .small}

**Concept**: numeric data visualization reveals biological patterns and data integrity within a real biomedical measure

**Dataset**: Serum creatinine values (mg/dL) from 500 patients — a direct numeric indicator of renal filtration

**Visualization**: Histogram showing the distribution of values across the patient population
:::
::: column

```{r}
#| fig-width: 8
#| fig-height: 9

set.seed(123)
creatinine <- c(rnorm(450, mean = 1.0, sd = 0.25),   # typical renal function
                rnorm(40, mean = 2.2, sd = 0.4),     # impaired subgroup
                rnorm(10, mean = 5.5, sd = 0.5))     # extreme outliers

library(ggplot2)

ggplot(data.frame(creatinine), aes(x = creatinine)) +
  geom_histogram(binwidth = 0.2, fill = "#2c7bb6", color = "white", boundary = 0) +
  labs(
    title = "Serum Creatinine Levels Across 500 Patients",
    x = "Serum Creatinine (mg/dL)",
    y = "Number of Patients"
  ) +
  theme_minimal(base_size = 14)
```
:::
:::


<style>
.tiny-table table { font-size: 0.8em; }        /* adjust as needed (0.8–0.9em) */
.tiny-table figcaption, .tiny-table .table-caption { font-size: 0.75em; }
</style>


<style>
.tiny-table table { font-size: 0.8em; }        /* adjust as needed (0.8–0.9em) */
.tiny-table figcaption, .tiny-table .table-caption { font-size: 0.75em; }
</style>

## <span style="font-size:80%">Example: Serum Creatinine Distribution</span> {.smaller}

::: columns
::: {.column .small}

**Histogram**:

- Central peak near 0.8–1.0 mg/dL → expected range for normal renal function

- Right-hand tail extending beyond 2.0 mg/dL → subgroup with possible renal impairment

- Extreme outliers (>5 mg/dL) → may represent dialysis patients, transcription errors, or unit inconsistencies

**Plot interpretation**:

- Even a simple numeric plot surfaces biological substructure (healthy vs. impaired) and data quality signals simultaneously

- Visualization transforms numbers into an interpretable clinical landscape — far more revealing than a table or summary statistic

:::
::: column

```{r}
#| fig-width: 8
#| fig-height: 9

set.seed(123)
creatinine <- c(rnorm(450, mean = 1.0, sd = 0.25),   # typical renal function
                rnorm(40, mean = 2.2, sd = 0.4),     # impaired subgroup
                rnorm(10, mean = 5.5, sd = 0.5))     # extreme outliers

library(ggplot2)

ggplot(data.frame(creatinine), aes(x = creatinine)) +
  geom_histogram(binwidth = 0.2, fill = "#2c7bb6", color = "white", boundary = 0) +
  labs(
    title = "Serum Creatinine Levels Across 500 Patients",
    x = "Serum Creatinine (mg/dL)",
    y = "Number of Patients"
  ) +
  theme_minimal(base_size = 14)
```
:::
:::


<style>
.tiny-table table { font-size: 0.8em; }        /* adjust as needed (0.8–0.9em) */
.tiny-table figcaption, .tiny-table .table-caption { font-size: 0.75em; }
</style>







<style>
.tiny-table table { font-size: 0.8em; }        /* adjust as needed (0.8–0.9em) */
.tiny-table figcaption, .tiny-table .table-caption { font-size: 0.75em; }
</style>


## <span style="font-size:80%">Summary</span> {.smaller}

::: columns
::: {.column width="0%"}

:::

::: {.column width="100%"}

- Numeric visualization transforms quantities into visible evidence

- Understand what numbers represent biologically before encoding them visually

- Maintain context and scale integrity — avoid misleading visuals

- Visualization is the first step of statistical modeling, not an afterthought
:::

:::
::: footer
:::















